---
title: "DSHB Capstone"
author: "Liam Smith-Becker, Andi Donnelly"
date: "2024-10-16"
output: html_document
---

```{r setup, include=FALSE}
library(httr)
library(readxl)
library(knitr)
library(tidyverse)
library(janitor)
library(psych)
library(dplyr)
library(janitor)
library(ggthemes)
library(ggplot2)
library(corrplot)
library(tidyr)
library(maps)
library(scales)
library(plotly)
library(gganimate)
library(broom)
library(Matrix)
library(lme4)
```



```{r}
df_2019 <- read.csv(gzfile("2019_details.csv.gz"))
df_2021 <- read.csv(gzfile("2021_details.csv.gz"))
df_2022 <- read.csv(gzfile("2022_details.csv.gz"))
df_2023 <- read.csv(gzfile("2023_details.csv.gz")) 
df_2024 <- read.csv(gzfile("2024_details.csv.gz")) 
```



```{r}
states_to_remove <- c("AMERICAN SAMOA", "DISTRICT OF COLUMBIA", "E PACIFIC", 
                      "GULF OF MEXICO", "HAWAII WATERS", "LAKE ERIE", 
                      "LAKE HURON", "LAKE MICHIGAN", "LAKE ONTARIO", 
                      "LAKE ST CLAIR", "LAKE SUPERIOR", "PUERTO RICO", "ATLANTIC NORTH", "ATLANTIC SOUTH", "GUAM", 
                      "ST LAWRENCE R","VIRGIN ISLANDS", "GULF OF ALASKA", "GUAM WATERS")

df_2019 <- df_2019[!df_2019$STATE %in% states_to_remove, ]
df_2021 <- df_2021[!df_2021$STATE %in% states_to_remove, ]
df_2022 <- df_2022[!df_2022$STATE %in% states_to_remove, ]
df_2023 <- df_2023[!df_2023$STATE %in% states_to_remove, ]
df_2024 <- df_2024[!df_2024$STATE %in% states_to_remove, ]
```


```{r}
df_2019 %>% janitor::clean_names()
df_2021 %>% janitor::clean_names()
df_2022 %>% janitor::clean_names()
df_2023 %>% janitor::clean_names()
df_2024 %>% janitor::clean_names()
```



```{r}
colnames(df_2019)
```

```{r}
table(df_2019$EVENT_TYPE)
table(df_2021$EVENT_TYPE)
table(df_2022$EVENT_TYPE)
table(df_2023$EVENT_TYPE)
table(df_2024$EVENT_TYPE)
```


```{r}
df_2019$EVENT_TYPE <- replace(df_2019$EVENT_TYPE, df_2019$EVENT_TYPE == "Hurricane", "Hurricane (Typhoon)")

df_2021$EVENT_TYPE <- replace(df_2021$EVENT_TYPE, df_2021$EVENT_TYPE == "Hurricane", "Hurricane (Typhoon)")

df_2024$EVENT_TYPE <- replace(df_2024$EVENT_TYPE, df_2024$EVENT_TYPE == "Hurricane", "Hurricane (Typhoon)")
```


```{r}
table(df_2019$STATE)
table(df_2021$STATE)
table(df_2022$STATE)
table(df_2023$STATE)
table(df_2024$STATE)
```


```{r}
stormfreq_2019 <- as.data.frame(table(df_2019$EVENT_TYPE))
colnames(stormfreq_2019) <- c("Value", "Frequency")
print(stormfreq_2019)

stormfreq_2021 <- as.data.frame(table(df_2021$EVENT_TYPE))
colnames(stormfreq_2021) <- c("Value", "Frequency")
print(stormfreq_2021)

stormfreq_2022 <- as.data.frame(table(df_2022$EVENT_TYPE))
colnames(stormfreq_2022) <- c("Value", "Frequency")
print(stormfreq_2022)

stormfreq_2023 <- as.data.frame(table(df_2023$EVENT_TYPE))
colnames(stormfreq_2023) <- c("Value", "Frequency")
print(stormfreq_2023)

stormfreq_2024 <- as.data.frame(table(df_2024$EVENT_TYPE))
colnames(stormfreq_2024) <- c("Value", "Frequency")
print(stormfreq_2024)
```



```{r}
statefreq_2019 <- as.data.frame(table(df_2019$STATE))
colnames(statefreq_2019) <- c("Value", "Frequency")
print(statefreq_2019)

statefreq_2021 <- as.data.frame(table(df_2021$STATE))
colnames(statefreq_2021) <- c("Value", "Frequency")
print(statefreq_2021)

statefreq_2022 <- as.data.frame(table(df_2022$STATE))
colnames(statefreq_2022) <- c("Value", "Frequency")
print(statefreq_2022)

statefreq_2023 <- as.data.frame(table(df_2023$STATE))
colnames(statefreq_2023) <- c("Value", "Frequency")
print(statefreq_2023)

statefreq_2024 <- as.data.frame(table(df_2024$STATE))
colnames(statefreq_2024) <- c("Value", "Frequency")
print(statefreq_2024)
```

```{r}
ggplot(stormfreq_2019, aes(x = Value, y = Frequency)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title="Storm Frequency Distribution 2019", x="Categories", y="Frequency")

ggplot(stormfreq_2021, aes(x = Value, y = Frequency)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title="Storm Frequency Distribution 2021", x="Categories", y="Frequency")

ggplot(stormfreq_2022, aes(x = Value, y = Frequency)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title="Storm Frequency Distribution 2022", x="Categories", y="Frequency")

ggplot(stormfreq_2023, aes(x = Value, y = Frequency)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title="Storm Frequency Distribution 2023", x="Categories", y="Frequency")

ggplot(stormfreq_2024, aes(x = Value, y = Frequency)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title="Storm Frequency Distribution 2024", x="Categories", y="Frequency")

```

```{r}
ggplot(statefreq_2019, aes(x = Value, y = Frequency)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title="State Frequency Distribution 2019", x="Categories", y="Frequency")

ggplot(statefreq_2021, aes(x = Value, y = Frequency)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title="State Frequency Distribution 2021", x="Categories", y="Frequency")

ggplot(statefreq_2022, aes(x = Value, y = Frequency)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title="State Frequency Distribution 2022", x="Categories", y="Frequency")

ggplot(statefreq_2023, aes(x = Value, y = Frequency)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title="State Frequency Distribution 2023", x="Categories", y="Frequency")

ggplot(statefreq_2024, aes(x = Value, y = Frequency)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title="State Frequency Distribution 2024", x="Categories", y="Frequency")
```


```{r}
stormfreq_2019
stormfreq_2021
stormfreq_2022
stormfreq_2023
stormfreq_2024
```

```{r}

storm_freq_dfs <- list(stormfreq_2019, stormfreq_2021, stormfreq_2022, stormfreq_2023, stormfreq_2024)

common_event_types <- Reduce(intersect, lapply(storm_freq_dfs, function(df) df$Value))

filter_common_types <- function(df, common_types) {
  df <- df[df$Value %in% common_types, ]
  df <- df[order(df$Value), ]  
  return(df)
}

stormfreq_2019 <- filter_common_types(stormfreq_2019, common_event_types)
stormfreq_2021 <- filter_common_types(stormfreq_2021, common_event_types)
stormfreq_2022 <- filter_common_types(stormfreq_2022, common_event_types)
stormfreq_2023 <- filter_common_types(stormfreq_2023, common_event_types)
stormfreq_2024 <- filter_common_types(stormfreq_2024, common_event_types)

```



```{r}
stormfreq_2019 <- stormfreq_2019 %>% rename(freq_2019 = Frequency)
stormfreq_2021 <- stormfreq_2021 %>% rename(freq_2021 = Frequency)
stormfreq_2022 <- stormfreq_2022 %>% rename(freq_2022 = Frequency)
stormfreq_2023 <- stormfreq_2023 %>% rename(freq_2023 = Frequency)
stormfreq_2024 <- stormfreq_2024 %>% rename(freq_2024 = Frequency)
stormfreq_2024
```




```{r}
storm_freq_dfs <- list(stormfreq_2019, stormfreq_2021, stormfreq_2022, stormfreq_2023, stormfreq_2024)
storm_freq <- Reduce(function(x, y) merge(x, y, by = "Value", all = TRUE), storm_freq_dfs)

storm_freq

storm_freq <- storm_freq %>% rename(event_type = Value)
```

```{r}

storm_freq_long <- storm_freq %>%
  pivot_longer(cols = starts_with("freq_"), names_to = "Year", values_to = "Frequency")


ggplot(storm_freq_long, aes(x = event_type, y = Frequency, fill = Year)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Frequency of Storm Events by Year", x = "Event Type", y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```
```{r}
sum(df_2024$EVENT_TYPE == "Hurricane (Typhoon)")
```


```{r}

ggplot(storm_freq_long, aes(x = Year, y = Frequency, color = event_type, group = event_type)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Frequency Trends of Storm Events Over Years", x = "Year", y = "Frequency") +
  theme_minimal() +
  theme(legend.position = "right")

```
```{r}

ggplot(storm_freq_long, aes(x = Year, y = event_type, fill = Frequency)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Heatmap of Storm Event Frequencies by Year", x = "Year", y = "Event Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```



```{r}
state_freq_combined <- bind_rows(
  mutate(statefreq_2019, Year = 2019),
  mutate(statefreq_2021, Year = 2021),
  mutate(statefreq_2022, Year = 2022),
  mutate(statefreq_2023, Year = 2023),
  mutate(statefreq_2024, Year = 2024)
)

class(state_freq_combined) 

head(state_freq_combined)
colnames(state_freq_combined)

```



```{r}
unique(state_freq_combined$Year) 
head(state_freq_combined$state)
storm_state_2023 <- filter(state_freq_combined, Year == 2023)

```



```{r}
us_states <- map_data("state")
state_freq_combined <- rename(state_freq_combined, state = Value)
```



```{r}
##2019

storm_state_2019 <- filter(state_freq_combined, Year == 2019)

storm_state_2019$state <- tolower(storm_state_2019$state)

map_data_2019 <- merge(us_states, storm_state_2019, by.x = "region", by.y = "state", all.x = TRUE)


ggplot(map_data_2019, aes(long, lat, group = group, fill = Frequency)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkgreen", na.value = "grey90") +
  labs(title = "Storm Event Frequency by State in 2019", fill = "Frequency", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())

##2021
storm_state_2021 <- filter(state_freq_combined, Year == 2021)

storm_state_2021$state <- tolower(storm_state_2021$state)

map_data_2021 <- merge(us_states, storm_state_2021, by.x = "region", by.y = "state", all.x = TRUE)


ggplot(map_data_2021, aes(long, lat, group = group, fill = Frequency)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkgreen", na.value = "grey90") +
  labs(title = "Storm Event Frequency by State in 2021", fill = "Frequency", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())

##2022
storm_state_2022 <- filter(state_freq_combined, Year == 2022)

storm_state_2022$state <- tolower(storm_state_2022$state)

map_data_2022 <- merge(us_states, storm_state_2022, by.x = "region", by.y = "state", all.x = TRUE)


ggplot(map_data_2022, aes(long, lat, group = group, fill = Frequency)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkgreen", na.value = "grey90") +
  labs(title = "Storm Event Frequency by State in 2022", fill = "Frequency", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())

##2023
storm_state_2023 <- filter(state_freq_combined, Year == 2023)

storm_state_2023$state <- tolower(storm_state_2023$state)

map_data_2023 <- merge(us_states, storm_state_2023, by.x = "region", by.y = "state", all.x = TRUE)

ggplot(map_data_2023, aes(long, lat, group = group, fill = Frequency)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkgreen", na.value = "grey90") +
  labs(title = "Storm Event Frequency by State in 2023", fill = "Frequency", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())

##2024
storm_state_2024 <- filter(state_freq_combined, Year == 2024)

storm_state_2024$state <- tolower(storm_state_2024$state)

map_data_2024 <- merge(us_states, storm_state_2024, by.x = "region", by.y = "state", all.x = TRUE)


ggplot(map_data_2024, aes(long, lat, group = group, fill = Frequency)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkgreen", na.value = "grey90") +
  labs(title = "Storm Event Frequency by State in 2024", fill = "Frequency", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())

```


```{r}
##2019
ggplot(storm_state_2019, aes(x = reorder(state, Frequency), y = Frequency, size = Frequency, color = Frequency)) +
  geom_point(alpha = 0.7) +
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Storm Event Frequency by State in 2019", x = "State", y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

##2021
ggplot(storm_state_2021, aes(x = reorder(state, Frequency), y = Frequency, size = Frequency, color = Frequency)) +
  geom_point(alpha = 0.7) +
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Storm Event Frequency by State in 2021", x = "State", y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

##2022
ggplot(storm_state_2022, aes(x = reorder(state, Frequency), y = Frequency, size = Frequency, color = Frequency)) +
  geom_point(alpha = 0.7) +
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Storm Event Frequency by State in 2022", x = "State", y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

##2023
ggplot(storm_state_2023, aes(x = reorder(state, Frequency), y = Frequency, size = Frequency, color = Frequency)) +
  geom_point(alpha = 0.7) +
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Storm Event Frequency by State in 2023", x = "State", y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

##2024
ggplot(storm_state_2024, aes(x = reorder(state, Frequency), y = Frequency, size = Frequency, color = Frequency)) +
  geom_point(alpha = 0.7) +
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Storm Event Frequency by State in 2024", x = "State", y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 


```
```{r}
ggplot(state_freq_combined, aes(x = Year, y = Frequency, fill = as.factor(Year))) +
  geom_bar(stat = "identity") +
  facet_wrap(~ state, scales = "free_y") +
  labs(title = "Storm Event Frequency Over Years by State", x = "Year", y = "Frequency") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
###this package is struggling a bit to change through - it might be the size of the data???

p <- ggplot(state_freq_combined, aes(x = Year, y = Frequency, fill = as.factor(Year))) +
  geom_bar(stat = "identity") +
  labs(title = "Storm Event Frequency Over Years by State", x = "Year", y = "Frequency") +
  theme_minimal() +
  theme(legend.position = "none")


p_interactive <- ggplotly(p) %>%
  layout(
    updatemenus = list(
      list(
        type = "dropdown",
        active = 0,
        buttons = lapply(unique(state_freq_combined$state), function(state) {
          list(
            method = "update",
            label = state,
            args = list(list(visible = state_freq_combined$state == state))
          )
        })
      )
    )
  )

p_interactive
```


```{r}
start_end <- state_freq_combined %>% filter(Year %in% c(2019, 2024))

ggplot(start_end, aes(x = factor(Year), y = Frequency, group = state, color = state)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Change in Storm Event Frequency (2019 vs. 2024)", x = "Year", y = "Frequency") +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r}
ggplot(state_freq_combined, aes(x = Year, y = Frequency, group = 1)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ state, scales = "free_y") +
  labs(title = "Storm Event Frequency Over Time by State", x = "Year", y = "Frequency") +
  theme_minimal()
```

```{r}
ggplot(state_freq_combined, aes(x = Year, y = reorder(state, Frequency), fill = Frequency)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Storm Event Frequency by State and Year", x = "Year", y = "State") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
state_freq_dfs <- list(statefreq_2019, statefreq_2021, statefreq_2022, statefreq_2023, statefreq_2024)
state_freq <- Reduce(function(x, y) merge(x, y, by = "Value", all = TRUE), state_freq_dfs)

state_freq

state_freq <- statefreq_2019 %>% rename(freq_2019 = Frequency)
state_freq

state_freq <- state_freq %>%
  full_join(statefreq_2021 %>% rename(freq_2021 = Frequency), by = "Value") %>%
  full_join(statefreq_2022 %>% rename(freq_2022 = Frequency), by = "Value") %>%
  full_join(statefreq_2023 %>% rename(freq_2023 = Frequency), by = "Value") %>%
  full_join(statefreq_2024 %>% rename(freq_2024 = Frequency), by = "Value")

state_freq

state_freq <- state_freq %>% rename(state = Value)
head(state_freq)
```

```{r}
state_freq_long <- state_freq %>%
  pivot_longer(
    cols = starts_with("freq_"),
    names_to = "Year",
    names_prefix = "freq_",
    values_to = "Frequency"
  ) %>%
  mutate(Year = as.integer(Year))
```

```{r}
lm_state <- state_freq_long %>%
  group_by(state) %>%
  summarize(model = list(lm(Frequency ~ Year, data = pick(everything())))) %>%
  mutate(tidy = purrr::map(model, broom::tidy)) %>%
  unnest(tidy)

print(lm_state)
```


```{r}
pm_state <- state_freq_long %>%
  group_by(state) %>%
  summarize(model = list(lm(Frequency ~ poly(Year, 2), data = pick(everything())))) %>%
  mutate(tidy = purrr::map(model, broom::tidy)) %>%
  unnest(tidy)

print(pm_state)

```


```{r}
filter(lm_state, state == "ALABAMA")
filter(pm_state, state == "ALABAMA")
```


```{r}
ranked_state <- lm_state %>%
  filter(term == "Year") %>%
  arrange(desc(estimate))  
ranked_state
```


###
A state-wise linear regression analysis was conducted to investigate frequency trends over time from 2019 to 2024. Each state's data was analyzed using Year as the predictor variable and Frequency as the outcome. The results revealed variations in trends across states. Texas demonstrated a significant upward trend, with frequency increasing by approximately 598.72 units annually (p ≈ 0.08), suggesting that specific factors might be contributing to this consistent rise. Oklahoma and Georgia showed positive, though less pronounced, trends with annual increases of 399.36 and 353.32 units, respectively. These positive trends in frequency imply potential growing activity or incidence rates in these states over the study period.

Some states exhibited a significant decline in frequency over time. Massachusetts and Pennsylvania both showed negative trends, with frequency decreasing by approximately -76.81 and -154.47 units per year, respectively. This suggests stabilizing or diminishing activity levels in these states, with Massachusetts's trend reaching statistical significance (p ≈ 0.05), while Pennsylvania’s trend was just shy of conventional significance levels. Alaska showed an upward trend, with a yearly increase of around 54.66 units (p = 0.02), indicating a steady rise in frequency in recent years.


```{r}
ggplot(state_freq_long, aes(x = Year, y = Frequency, color = state)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  labs(title = "Linear Trend of Frequency Over Time by State", y = "Frequency") +
  theme_minimal()

###hate this figure deeply
```

```{r}
ggplot(state_freq_long, aes(x = Year, y = Frequency, color = state)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE) +
  labs(title = "Polynomial Trend of Frequency Over Time by State", y = "Frequency") +
  theme_minimal()

###not all that important and i don't think we should even mention this.
```



```{r}
lmer_storm <- lmer(Frequency ~ Year + (1 | event_type), data = storm_freq_long)

summary(lmer_storm)
```



##
A linear mixed model was conducted to examine how storm frequency varies by year and storm type. The model included year as a fixed effect and storm as a random effect to account for differences in baseline frequencies across storm types. Results show that storm types had significant baseline variability, with a random intercept standard deviation of 3558. This variance suggests that certain storm types tend to occur more frequently than others, independent of yearly changes. The fixed effects for year did not have significant changes in frequency over time. Yearly effects varied but didn’t reach statistical significance, with estimates ranging from -184.47 to 338.83 and associated t-values below 1.96. The lack of significant change suggests that frequency doesn’t follow a clear upward or downward trend across the years.

Overall, the model shows that while different storm types vary in baseline frequency, there’s no consistent trend in frequency across years. This implies that storm frequency remains relatively stable over time for most storms. Ultimately, there’s importance of storm-specific baselines rather than year-over-year changes for overall storm frequency patterns.

########################## REMOVING TEXAS ##############################

```{r}
state_freq_rmtexas_long <- state_freq_long %>% filter(state != "TEXAS")
```

```{r}
ggplot(state_freq_rmtexas_long, aes(x = Year, y = state, fill = Frequency)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Heatmap of Storm Event Frequencies W/O Texas by Year", x = "Year", y = "Event Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

```{r}
start_end_1 <- state_freq_rmtexas_long %>% filter(Year %in% c(2019, 2024))

ggplot(start_end_1, aes(x = factor(Year), y = Frequency, group = state, color = state)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Change in Storm Event Frequency Without Texas (2019 vs. 2024)", x = "Year", y = "Frequency") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
state_freq_combined_rmtx <- state_freq_combined %>% filter(state != "TEXAS")
```


```{r}
##2019

rmtxstorm_state_2019 <- filter(state_freq_combined_rmtx, Year == 2019)

rmtxstorm_state_2019$state <- tolower(rmtxstorm_state_2019$state)

map_data_2019 <- merge(us_states, rmtxstorm_state_2019, by.x = "region", by.y = "state", all.x = TRUE)


ggplot(map_data_2019, aes(long, lat, group = group, fill = Frequency)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkgreen", na.value = "grey90") +
  labs(title = "Storm Event Frequency by State in 2019", fill = "Frequency", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())

##2021
rmtxstorm_state_2021 <- filter(state_freq_combined_rmtx,, Year == 2021)

rmtxstorm_state_2021$state <- tolower(rmtxstorm_state_2021$state)

map_data_2021 <- merge(us_states, rmtxstorm_state_2021, by.x = "region", by.y = "state", all.x = TRUE)


ggplot(map_data_2021, aes(long, lat, group = group, fill = Frequency)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkgreen", na.value = "grey90") +
  labs(title = "Storm Event Frequency by State in 2021", fill = "Frequency", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())

##2022
rmtxstorm_state_2022 <- filter(state_freq_combined_rmtx,, Year == 2022)

rmtxstorm_state_2022$state <- tolower(rmtxstorm_state_2022$state)

map_data_2022 <- merge(us_states, rmtxstorm_state_2022, by.x = "region", by.y = "state", all.x = TRUE)


ggplot(map_data_2022, aes(long, lat, group = group, fill = Frequency)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkgreen", na.value = "grey90") +
  labs(title = "Storm Event Frequency by State in 2022", fill = "Frequency", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())

##2023
rmtxstorm_state_2023 <- filter(state_freq_combined_rmtx, Year == 2023)

rmtxstorm_state_2023$state <- tolower(rmtxstorm_state_2023$state)

map_data_2023 <- merge(us_states, rmtxstorm_state_2023, by.x = "region", by.y = "state", all.x = TRUE)

ggplot(map_data_2023, aes(long, lat, group = group, fill = Frequency)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkgreen", na.value = "grey90") +
  labs(title = "Storm Event Frequency by State in 2023", fill = "Frequency", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())

##2024
rmtxstorm_state_2024 <- filter(state_freq_combined_rmtx,, Year == 2024)

rmtxstorm_state_2024$state <- tolower(rmtxstorm_state_2024$state)

map_data_2024 <- merge(us_states, rmtxstorm_state_2024, by.x = "region", by.y = "state", all.x = TRUE)


ggplot(map_data_2024, aes(long, lat, group = group, fill = Frequency)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkgreen", na.value = "grey90") +
  labs(title = "Storm Event Frequency by State in 2024", fill = "Frequency", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), panel.grid = element_blank())
```

```{r}
txstates_to_remove <- c("AMERICAN SAMOA", "DISTRICT OF COLUMBIA", "E PACIFIC", 
                      "GULF OF MEXICO", "HAWAII WATERS", "LAKE ERIE", 
                      "LAKE HURON", "LAKE MICHIGAN", "LAKE ONTARIO", 
                      "LAKE ST CLAIR", "LAKE SUPERIOR", "PUERTO RICO", "ATLANTIC NORTH", "ATLANTIC SOUTH", "GUAM", 
                      "ST LAWRENCE R","VIRGIN ISLANDS", "GULF OF ALASKA", "GUAM WATERS", "TEXAS")

txdf_2019 <- df_2019[!df_2019$STATE %in% txstates_to_remove, ]
txdf_2021 <- df_2021[!df_2021$STATE %in% txstates_to_remove, ]
txdf_2022 <- df_2022[!df_2022$STATE %in% txstates_to_remove, ]
txdf_2023 <- df_2023[!df_2023$STATE %in% txstates_to_remove, ]
txdf_2024 <- df_2024[!df_2024$STATE %in% txstates_to_remove, ]
```


```{r}
txdf_2019 %>% janitor::clean_names()
txdf_2021 %>% janitor::clean_names()
txdf_2022 %>% janitor::clean_names()
txdf_2023 %>% janitor::clean_names()
txdf_2024 %>% janitor::clean_names()
```


```{r}
txdf_2019$EVENT_TYPE <- replace(txdf_2019$EVENT_TYPE, txdf_2019$EVENT_TYPE == "Hurricane", "Hurricane (Typhoon)")

txdf_2021$EVENT_TYPE <- replace(txdf_2021$EVENT_TYPE, txdf_2021$EVENT_TYPE == "Hurricane", "Hurricane (Typhoon)")

txdf_2024$EVENT_TYPE <- replace(txdf_2024$EVENT_TYPE, txdf_2024$EVENT_TYPE == "Hurricane", "Hurricane (Typhoon)")
```


```{r}
rmtxstormfreq_2019 <- as.data.frame(table(txdf_2019$EVENT_TYPE))
colnames(rmtxstormfreq_2019) <- c("Value", "Frequency")
print(rmtxstormfreq_2019)

rmtxstormfreq_2021 <- as.data.frame(table(txdf_2021$EVENT_TYPE))
colnames(rmtxstormfreq_2021) <- c("Value", "Frequency")
print(rmtxstormfreq_2021)

rmtxstormfreq_2022 <- as.data.frame(table(txdf_2022$EVENT_TYPE))
colnames(rmtxstormfreq_2022) <- c("Value", "Frequency")
print(rmtxstormfreq_2022)

rmtxstormfreq_2023 <- as.data.frame(table(txdf_2023$EVENT_TYPE))
colnames(rmtxstormfreq_2023) <- c("Value", "Frequency")
print(rmtxstormfreq_2023)

rmtxstormfreq_2024 <- as.data.frame(table(txdf_2024$EVENT_TYPE))
colnames(rmtxstormfreq_2024) <- c("Value", "Frequency")
print(rmtxstormfreq_2024)
```

```{r}
rmtxstorm_freq_dfs <- list(rmtxstormfreq_2019, rmtxstormfreq_2021, rmtxstormfreq_2022, rmtxstormfreq_2023, rmtxstormfreq_2024)
rmtxstorm_freq <- Reduce(function(x, y) merge(x, y, by = "Value", all = TRUE), rmtxstorm_freq_dfs)

rmtxstorm_freq

rmtxstorm_freq <- rmtxstormfreq_2019 %>% rename(rmtxfreq_2019 = Frequency)
rmtxstorm_freq

rmtxstorm_freq <- rmtxstorm_freq %>%
  full_join(rmtxstormfreq_2021 %>% rename(rmtxfreq_2021 = Frequency), by = "Value") %>%
  full_join(rmtxstormfreq_2022 %>% rename(rmtxfreq_2022 = Frequency), by = "Value") %>%
  full_join(rmtxstormfreq_2023 %>% rename(rmtxfreq_2023 = Frequency), by = "Value") %>%
  full_join(rmtxstormfreq_2024 %>% rename(rmtxfreq_2024 = Frequency), by = "Value")

rmtxstorm_freq

rmtxstorm_freq <- rmtxstorm_freq %>% rename(event_type = Value)
head(rmtxstorm_freq)
```


```{r}
rmtxstorm_freq_long <- rmtxstorm_freq %>%
  pivot_longer(cols = starts_with("rmtxfreq_"), names_to = "Year", values_to = "Frequency")
```


```{r}
ggplot(rmtxstorm_freq_long, aes(x = Year, y = Frequency, color = event_type, group = event_type)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Frequency Trends of Storm Events W/O Texas Over Years", x = "Year", y = "Frequency") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
rmtxstate_freq <- state_freq %>% filter(state != "TEXAS")
```

```{r}

state_freq_rmtexas_long <- rmtxstate_freq %>%
  pivot_longer(
    cols = starts_with("freq_"),
    names_to = "Year",
    names_prefix = "freq_",
    values_to = "Frequency"
  ) %>%
  mutate(Year = as.integer(Year))
```

```{r}
lm_rmtxstate <- state_freq_rmtexas_long %>%
  group_by(state) %>%
  summarize(model = list(lm(Frequency ~ Year, data = pick(everything())))) %>%
  mutate(tidy = purrr::map(model, broom::tidy)) %>%
  unnest(tidy)

print(lm_rmtxstate)
```

A state-wise linear regression analysis was done to look at frequency trends from 2019 to 2024. Year was used as the predictor variable and Frequency as the outcome. Trends varied across states. Alabama showed a positive trend, with frequency increasing by about 263.65 units per year (p ≈ 0.06), while Georgia and Alaska also had upward trends, with annual increases of 353.32 (p ≈ 0.06) and 54.66 units (p = 0.02), respectively. These increases hint at rising activity or incidence in these states. Some states showed declines. Massachusetts had a significant negative trend, with frequency dropping by about -76.81 units yearly (p ≈ 0.05), and Pennsylvania also showed a decrease of -154.47 units per year (p ≈ 0.05), though not quite at significance. Florida showed an upward trend, with frequency growing by roughly 151.45 units per year (p ≈ 0.04), suggesting a steady rise in recent years.


```{r}
ggplot(state_freq_rmtexas_long, aes(x = Year, y = Frequency, color = state)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  labs(title = "Linear Trend of Frequency Over Time by State", y = "Frequency") +
  theme_minimal()
```


```{r}
lmer_rmtxstorm <- lmer(Frequency ~ Year + (1 | event_type), data = rmtxstorm_freq_long)

summary(lmer_rmtxstorm)
```

A linear mixed model was run to check how storm frequency varies by year and storm type without Texas. The model used year as a fixed effect and storm type as a random effect to find differences in baseline frequency across storm types. Results showed significant baseline variability in storm types, with a random intercept standard deviation of 3048.5, indicating that certain storm types naturally occur more often than others, regardless of year. The fixed effects for year didn’t show significant changes in frequency over time. Yearly estimates ranged from -181.85 to 272.52, with t-values below 1.96, meaning frequency didn’t follow a clear trend up or down across years. The model shows that while storm types differ in baseline frequency, storm frequency itself remains pretty stable year to year. This suggests the importance of storm-specific baselines over any yearly changes.
